The Astropy Project: \\ Sustaining and Growing a Community-oriented Open-source Project and the Latest Major Release (v5.0) of the Core Package

### 2022-04-24 src/ms.tex
rephrase around mixin
Signed-off-by: nstarman <nstarkman@protonmail.com>

\paragraph{Table Mixin Columns}

Within \astropypkg and \python, there are numerous ways to represent and store
array-valued data. Some of the differences are historical:
\astropyapidoc{table.Column}{table.Column} and
\astropyapidoc{io.fits.Column}{io.fits.Column} are not the same. Some
differences are computational: \package{numpy}, \package{cupy}, and
\package{dask} arrays have almost identical APIs, but are optimized for
different use cases. Lastly, some differences are inherent: \astropyQuantity,
\astropyTime, and \astropySkyCoord\ represent fundamentally different types of
objects.

We aim to make it possible for any array-valued data to be used as a column in a
table. A well-defined protocol for mixin-columns has been developed for
\astropytable, allowing the original object to be used as a column with a
familiar API, and to ``round-trip'' through tables with no loss of data or
attributes. With this protocol, it is now possible to store \astropy native
objects -- including \astropyTime, \astropyQuantity, and \astropySkyCoord\ --
within a \astropyTable\ and write these to various file formats, such as FITS.
For objects not already covered by the mixin protocol, functions can be
registered with \astropyTable\ to convert any array-like object into a mixin
column. As an example, the mixin functions are used to integrate \package{dask}
arrays with \astropyTable, allowing cloud-stored or cluster-scale data to be
used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to
other types of coordinates in WCS, time requires more metadata: format, scale,
position, reference, etc. This metadata had to be manually specified and the
nuances understood by the user. \astropyfits could read this data as a standard
\astropyFitsColumn\, but would not interpret the time-related metadata and
attributes. Through the support of the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{To learn more about this project,
see the final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a mixin column. For
backward compatibility, this feature may be turned off.

@@

As mentioned previously, \package{dask} arrays may be used as a mixin column in
\astropyTable. Now, \package{dask} may also be a data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only
computed while written, avoiding excessive memory use. Furthermore, \astropyTable\ can
be appended to an existing FITS file, where \package{dask} mixin columns can
interoperate seamlessly between them.

>>>>

\paragraph{Table Column Types}

Within \astropypkg and \python, there are numerous ways to represent and store
array-valued data. Some of the differences are historical: \astropytable and
\astropyapidoc{io.fits.BinTableHDU}{io.fits.BinTableHDU} are not the same since
\astropyfits was originally developed in a separate package (\package{PyFITS})
yet both are for holding tabular data. Some differences are computational:
\package{numpy}, \package{cupy}, and \package{dask} arrays have almost identical
APIs, but are optimized for different use cases. Lastly, some differences are
inherent: \astropyQuantity, \astropyTime, and \astropySkyCoord\ represent
fundamentally different types of objects.

\astropytable is \astropypkg's most general container for array-valued data.
Rather than requiring component data be converted to a
\astropyapidoc{table.Column}{table.Column}, a well-defined protocol has been
developed for any array-valued data to be used in a \astropytable, without
having to convert the data. The protocol for ``mixing'' column types provides
the same familiar table API, yet loses none the object's data or attributes.
With this protocol, it is now possible to store \astropy native objects --
including \astropyTime, \astropyQuantity, and \astropySkyCoord\ -- within a
\astropyTable\ and write these to various file formats, such as FITS. For
objects not already covered by the ``mixing'' protocol, functions can be
registered with \astropyTable\ to enable the interoperability. As an example,
\astropypkg arelady implements the ``mixing'' functions to integrate
\package{dask} arrays with \astropyTable, allowing cloud-stored or cluster-scale
data to be used as a column in a \astropyTable.

@@

The FITS standard was extended to rigorously describe time coordinates in the
WCS framework \citep{FITS-Time:2015}. Compared to other types of coordinates in
WCS, time requires more metadata: format, scale, position, reference, etc. This
metadata had to be manually specified and the nuances understood by the user.
\astropyfits could read this data as a standard \astropyFitsColumn, but would
not interpret the time-related metadata and attributes. Through the support of
the
\href{https://summerofcode.withgoogle.com/archive/2017/projects/4778482366152704}{Google
Summer of Code} 2017 program,\footnote{To learn more about this project, see the
final report
\href{https://aaryapatil.wordpress.com/2017/08/28/a-mixin-protocol-for-seamless-interoperability/}{A
mixin protocol for seamless interoperability}.} the \astropyfits package now
interprets time data correctly, using \astropyTime\ as a column. For backward
compatibility, this feature may be turned off.

@@

As mentioned previously, \package{dask} arrays may be used as a column in
\astropyTable. Now, \package{dask} may also be a data array in
\astropyapidoc{io.fits.PrimaryHDU}{FITS HDU} and only computed while written,
avoiding excessive memory use. Furthermore, \astropyTable\ can be appended to an
existing FITS file, where \package{dask} columns can interoperate seamlessly
between them.
